The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-15 manuscript/manuscript.Rnw
comment out OU and spider monkey sections

We used the software \ac{OUCH} in the R statistical environment to
investigate significant shifts in gene expression levels over
time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
institutes an \ac{OU} model of evolution, which allows for different
optima for quantitative trait values along user-specified branches of
a phylogenetic tree. The \ac{OU} model explicitly accounts for change
in quantitative trait values by introducing two parameters into the
model of stochastic Brownian motion of change in trait value along a
phylogenetic tree through time. These parameters are $\theta$, an
optimum trait value, and $\alpha$, the strength of selection
(interpreted as “pull” toward the optimum). In addition, the \ac{OU}
model can also integrate evolutionary change by assuming that each
user-specified set of braches has its own separate \ac{OU} parameters
\cite{Butler.2004:phlogeneticcomparativeanalysis}.

\DLA{The description of the OU process is terrible; it needs to be
  completely redone}

Various \ac{OU} branch models were tested against a model in which no
variation in optima ($\theta$) among branches was allowed. The
following branches were specified in separate models to have a
different optimum than all other branches: human, bonobo, mouse, dog,
cow, elephant, opossum, human+bonobo, Euarchontoglires, and
Boreoeutheria. Likelihood values were computed for each branch model
and compared to an \ac{OU} model with one optimum using likelihood
ratio tests with df=1. We filtered out genes with universally low
expression levels, and included only genes in which expression in at
least one species had an $\mathrm{FPKM}\ge10$ as described above. We
corrected for multiple testing using the \ac{FDR} and using the
Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
used the time-calibrated phylogenetic tree of
\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
not used in this analyses pruned.

We also calculated the enrichment of functional categories in the
lists of genes which displayed a significant shift in expression on
each branch using the same methodology as in the per-species gene
lists \DLA{reference the tables once this analysis is complete}.

@@

\subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}

To verify expression levels in human and bonobo, we also generated
RNA-Seq data for the placenta transcriptome of the black-headed spider
monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
Ateles is outside the clade, which includes hominids and can be used
to polarize expression values of highly expressed genes in Homo and
Pan. We sampled placental tissue from an Ateles placenta expelled from
the mother shortly after birth (see Supplemental Table X). We used
similar procedures discussed above to extract and sequence this
sample. [DEREK—I’m unsure how to proceed on describing this so as to
not step on Natalie’s toes]

\DLA{Complete discussion of the ateles placenta}

>>>>

% We used the software \ac{OUCH} in the R statistical environment to
% investigate significant shifts in gene expression levels over
% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
% institutes an \ac{OU} model of evolution, which allows for different
% optima for quantitative trait values along user-specified branches of
% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
% in quantitative trait values by introducing two parameters into the
% model of stochastic Brownian motion of change in trait value along a
% phylogenetic tree through time. These parameters are $\theta$, an
% optimum trait value, and $\alpha$, the strength of selection
% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
% model can also integrate evolutionary change by assuming that each
% user-specified set of braches has its own separate \ac{OU} parameters
% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
% 
% \DLA{The description of the OU process is terrible; it needs to be
%   completely redone}
% 
% Various \ac{OU} branch models were tested against a model in which no
% variation in optima ($\theta$) among branches was allowed. The
% following branches were specified in separate models to have a
% different optimum than all other branches: human, bonobo, mouse, dog,
% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
% Boreoeutheria. Likelihood values were computed for each branch model
% and compared to an \ac{OU} model with one optimum using likelihood
% ratio tests with df=1. We filtered out genes with universally low
% expression levels, and included only genes in which expression in at
% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
% corrected for multiple testing using the \ac{FDR} and using the
% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
% used the time-calibrated phylogenetic tree of
% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
% not used in this analyses pruned.
% 
% We also calculated the enrichment of functional categories in the
% lists of genes which displayed a significant shift in expression on
% each branch using the same methodology as in the per-species gene
% lists \DLA{reference the tables once this analysis is complete}.

@@

% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
% 
% To verify expression levels in human and bonobo, we also generated
% RNA-Seq data for the placenta transcriptome of the black-headed spider
% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
% Ateles is outside the clade, which includes hominids and can be used
% to polarize expression values of highly expressed genes in Homo and
% Pan. We sampled placental tissue from an Ateles placenta expelled from
% the mother shortly after birth (see Supplemental Table X). We used
% similar procedures discussed above to extract and sequence this
% sample. [DEREK—I’m unsure how to proceed on describing this so as to
% not step on Natalie’s toes]
% 
% \DLA{Complete discussion of the ateles placenta}
